---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Trends in heavy metal concentrations in *Mytilus edulis* from several European nations"
subtitle: "https://github.com/slm119/ENV872_FinalProject"
author: "Sena McCrory"
fontsize: 12pt
mainfont: Times New Roman
editor_options: 
  chunk_output_type: console
---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
library(tidyverse)
library(scales)
library(nationalparkcolors)
library(viridis)

# load data sets
mytilus.metals <- read.csv("./Data/Processed/ICES_Mytilusedulis_Cd_Pb.csv")
mytilus.metals$DATE <- as.Date(mytilus.metals$DATE, format = "%d/%m/%Y")

mytilus.pb.monthly_post1990.interp <- read.csv("./Data/Processed/pb_monthly_post1990_interp.csv")
mytilus.cd.monthly_post1990.interp <- read.csv("./Data/Processed/cd_monthly_post1990_interp.csv")
mytilus.pb.monthly_post1990.interp$Month.Year <- as.Date(mytilus.pb.monthly_post1990.interp$Month.Year, format = "%Y-%m-%d")
mytilus.cd.monthly_post1990.interp$Month.Year <- as.Date(mytilus.cd.monthly_post1990.interp$Month.Year, format = "%Y-%m-%d")

# aggregate by yearly median 1979 - 2018
yearly.data <- mytilus.metals %>%
   group_by(MYEAR, PARAM)%>%
   dplyr::summarise(Conc = median(Value.mgperkg))

#subset post 1990
mytilus.post1990 <- mytilus.metals%>%
   filter(MYEAR >= 1990)%>%
   droplevels()

#change Country level order
mytilus.post1990$Country <- factor(mytilus.post1990$Country, levels = 
                                    c("Norway", "United Kingdom", 
                                      "Ireland", "Denmark", "The Netherlands",
                                      "Germany", "Spain", "Belgium", "Iceland", 
                                      "Poland", "Lithuania"))

# Set ggplot theme
my.theme <- theme_minimal(base_size = 14)+
  theme(legend.position = "top")
theme_set(my.theme)
my.colors <- park_palette("CraterLake")[c(5,7,1,2)]
```


# Rationale and Research Questions



\newpage

# Dataset Information



\newpage

# Exploratory Analysis 

The ICES database includes *Mytilus edulis* sample data from 1979 to 2018 for X European countries. Most samples were collected near shorelines and within some enclosed bay or lake areas (Fig 1).

```{r, echo=FALSE, fig.cap="Yearly median concentrations of cadmium and lead in *Mytilus edulis* ICES monitoring data from 1979 to 2018."}
# yearly plot
ggplot(yearly.data, aes(x=MYEAR, y = Conc, color = PARAM))+
   geom_line(size = .8)+
   geom_point()+
   scale_color_manual(values = my.colors)+
   ylim(0,1)+
   facet_wrap(~PARAM, nrow = 2, scales = "fixed")+
   theme(legend.position = "none")+
   labs(x = "Year", y = "Conc (mg metal/kg wet wt)")

```


```{r , echo=FALSE, fig.cap="Sample locations from 1990 to 2018", out.width = '90%'}
#knitr::include_graphics("./Output/StudyRegionMap.pdf")
```


```{r, echo=FALSE, fig.cap="Concentrations of metals in *Mytilus edulis* over time"}
# all sample conc 1990 - 2018 - excluding one outlier in Cd for visualiztion purposed
ggplot(mytilus.post1990 %>%
          filter((PARAM == "Cd" & Value.mgperkg < 20) | PARAM == "Pb"), aes(x=as.Date(DATE), y = Value.mgperkg))+
   geom_point(aes(color = PARAM), alpha = .5)+
   scale_color_manual(values = my.colors)+
   #scale_y_log10(labels = trans_format("log10", math_format(10^.x)))+
   labs(x= "Date", y = "Conc (mg metal/kg wet wt)", color = "Metal")+
   facet_wrap(~PARAM, scales = "free_y", ncol = 1)+
   theme(legend.position = "none")
```


```{r, echo=FALSE, fig.cap="Concentrations of metals in *Mytilus edulis* by country"}
# #histogram
ggplot(mytilus.post1990, aes(x=Value.mgperkg, fill = PARAM))+
   geom_histogram(bins = 40)+
   scale_x_log10(labels = trans_format("log10", math_format(10^.x)))+
   scale_fill_manual(values = my.colors)+
   facet_wrap(~PARAM, scales = "fixed")+
   labs(x="Log Concentration (mg/kg wet wt)", y = "Count", fill = "Metal")+
   theme(legend.position = "none")
```


```{r, echo=FALSE, fig.cap="Sample count by country"}
# violin by country
ggplot(mytilus.post1990, aes(x=Country, y = Value.mgperkg, fill = PARAM))+
   geom_violin(position = "dodge", scale = "count", draw_quantiles = .5, size = 1)+
   scale_fill_manual(values = my.colors)+
   scale_y_log10(labels = trans_format("log10", math_format(10^.x)))+
   labs(fill = "Metal")+
   coord_flip()+
   facet_wrap(~PARAM)+
   theme(legend.position = "none")+
   labs(y="Conc (mg/kg wet wt)", x = "")
```


\newpage

# Analysis



## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
