---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "Insert title of project here"
subtitle: "Web address for GitHub repository"
author: "Sena McCrory"
fontsize: 12pt
mainfont: Times New Roman
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages

library(tidyverse)
library(viridis)
library(lubridate)


# Set your ggplot theme
my.theme <- theme_minimal()+
  theme(legend.position = "top")
theme_set(my.theme)

# Load your datasets
ICES_metals.in.biota <- read.csv("./Data/Raw/ContaminantsBiota20202263412jpufqdsst5td1sxozv2woyiz.csv")
table(ICES_metals.in.biota$PARAM)
table(ICES_metals.in.biota$Country)

my.metals <- c("HG", "CD", "PB", "AS", "CR", "CR3+", "CR6+", "CU", "NI", "MN", "FE", "BA")
ICES_my.metals <- ICES_metals.in.biota %>%
  filter(PARAM %in% my.metals)%>%
  droplevels()
table(ICES_my.metals$PARAM)
table(ICES_my.metals$Country)
table(ICES_my.metals$Species)

species.summary <- ICES_my.metals %>%
  group_by(Species)%>%
  tally()
summary(species.summary$n)
top.5.species <- c("Mytilus edulis", "Gadus morhua", "Platichthys flesus", "Clupea harengus", "Limanda limanda")

ices_metals.top5 <- ICES_my.metals%>%
  filter(Species %in% top.5.species)%>%
  filter(MUNIT == "ug/kg")%>%
  droplevels()
#table(ices_metals.top5$MUNIT)
#ices_metals.top5$MYEAR <- as.Date(ices_metals.top5$MYEAR, format = "%Y", origin = "1970-01-01")

ggplot(ices_metals.top5, aes(x=MYEAR, y = Value))+
  geom_point(aes(color = Species), alpha = .5, size = 2, shape = 1)+
  facet_wrap(~PARAM, scales = "free")+
  scale_color_viridis_d(end = .9, option = "magma")
# barium was only measured recently, may be trend for As, difficult to tell for others due to some high outliers

ices_as <- ices_metals.top5%>%
  filter(PARAM == "AS")%>%
  droplevels()
ggplot(ices_as, aes(x=MYEAR, y = Value))+
  geom_point(aes(color = Species), alpha = .5, size = 2, shape = 1)+
  #ylim(c(0,10000))+
  scale_color_viridis_d(end = .9, option = "magma")+
  facet_wrap(~Species, scales = "free_y")
# gap in clupea harengus monitoring

#Barium
ices_ba <- ices_metals.top5%>%
  filter(PARAM == "BA")%>%
  droplevels()
ggplot(ices_ba, aes(x=MYEAR, y = Value))+
  geom_point(aes(color = Species), alpha = .5, size = 2, shape = 1)+
  #ylim(c(0,10000))+
  scale_color_viridis_d(end = .9, option = "magma")+
  facet_wrap(~Species, scales = "free_y")
# really sparse, eliminate from datasets

#Cadmium
ices_cd <- ices_metals.top5%>%
  filter(PARAM == "CD")%>%
  droplevels()
ggplot(ices_cd, aes(x=MYEAR, y = Value))+
  geom_point(aes(color = Species), alpha = .5, size = 2, shape = 1)+
  #ylim(c(0,10000))+
  scale_color_viridis_d(end = .9, option = "magma")+
  facet_wrap(~Species, scales = "free_y")
#outliers all over the place

#Chromium
ices_cr <- ices_metals.top5%>%
  filter(PARAM == "CR")%>%
  droplevels()
ggplot(ices_cr, aes(x=MYEAR, y = Value))+
  geom_point(aes(color = Species), alpha = .5, size = 2, shape = 1)+
  #ylim(c(0,10000))+
  scale_color_viridis_d(end = .9, option = "magma")+
  facet_wrap(~Species, scales = "free_y")
#good data for mytilus and not much else

# copper
ices_cu <- ices_metals.top5%>%
  filter(PARAM == "CU")%>%
  droplevels()
ggplot(ices_cu, aes(x=MYEAR, y = Value))+
  geom_point(aes(color = Species), alpha = .5, size = 2, shape = 1)+
  #ylim(c(0,10000))+
  scale_color_viridis_d(end = .9, option = "magma")+
  facet_wrap(~Species, scales = "free_y")
#outliers for mytilus

ices_hg <- ices_metals.top5%>%
  filter(PARAM == "HG")%>%
  droplevels()
ggplot(ices_hg, aes(x=MYEAR, y = Value))+
  geom_point(aes(color = Species), alpha = .5, size = 2, shape = 1)+
  #ylim(c(0,10000))+
  scale_color_viridis_d(end = .9, option = "magma")+
  facet_wrap(~Species, scales = "free_y")
# big outliers for mytilus and platichthys

ices_ni <- ices_metals.top5%>%
  filter(PARAM == "NI")%>%
  droplevels()
ggplot(ices_ni, aes(x=MYEAR, y = Value))+
  geom_point(aes(color = Species), alpha = .5, size = 2, shape = 1)+
  #ylim(c(0,10000))+
  scale_color_viridis_d(end = .9, option = "magma")+
  facet_wrap(~Species, scales = "free_y")
#lots of data gaps - maybe remove this one too

ices_pb <- ices_metals.top5%>%
  filter(PARAM == "PB")%>%
  droplevels()
ggplot(ices_pb, aes(x=MYEAR, y = Value))+
  geom_point(aes(color = Species), alpha = .5, size = 2, shape = 1)+
  #ylim(c(0,10000))+
  scale_color_viridis_d(end = .9, option = "magma")+
  facet_wrap(~Species, scales = "free_y")
# outliers for several species may be problematic? 

# perhaps just focus on mytils edulis? and maybe gadus morhua, too if time

```

review of metal concentrations and bioavailability in marine environments
https://scialert.net/fulltextmobile/?doi=jas.2004.1.20

mytilus edulis heavy metal concentrations https://www.researchgate.net/publication/14507180_Heavy_Metals_in_Blue_Mussels_Mytilus_edulis_in_the_Bergen_Harbor_Area_Western_Norway 

## Mytilus edulis only, metals to look at - Cd, Pb, Hg, Cu, Zn, remove "suspect" values from dataset
```{r}
ices_mytilus.edulis_cd.pb.hg.cu <- ices_metals.top5 %>%
  filter(Species == "Mytilus edulis" & 
           PARAM %in% c("CD", "PB", "HG", "CU"))%>%  
  droplevels()
ices_mytilus.edulis_cd.pb.hg.cu$DATE <- as.Date(ices_mytilus.edulis_cd.pb.hg.cu$DATE, format = "%d/%m/%Y")



table(ices_mytilus.edulis_cd.pb.hg.cu$VFLAG)
table(ices_mytilus.edulis_cd.pb.hg.cu$MATRX) # filer to include only soft body
table(ices_mytilus.edulis_cd.pb.hg.cu$BASIS) # dry weight or wet weight?? wet weight has more samples, but what about years and country coverage

wet <- ices_mytilus.edulis_cd.pb.hg.cu%>%
  filter(BASIS =="W")%>%
  droplevels()
summary(wet$Country)
dry <- ices_mytilus.edulis_cd.pb.hg.cu%>%
  filter(BASIS =="D")%>%
  droplevels()
summary(dry$Country)
#let's choose wet weight samples only just to make life simpler

ices_mytilus.edulis_cd.pb.hg.cu <- ices_metals.top5 %>%
  filter(Species == "Mytilus edulis" & 
           PARAM %in% c("CD", "PB", "HG", "CU", "ZN") & 
           VFLAG != "S" &  # remove suspect value measurments
           MATRX == "SB" & # soft body measurments only (many more than other matrices)
           BASIS == "W")%>%  # wet weight samples only (remove dry weight)
  droplevels()
ices_mytilus.edulis_cd.pb.hg.cu$DATE <- as.Date(ices_mytilus.edulis_cd.pb.hg.cu$DATE, format = "%d/%m/%Y")



ggplot(ices_mytilus.edulis_cd.pb.hg.cu, aes(x=as.Date(DATE), y = Value))+
  geom_point(aes(color = PARAM), shape = 1, alpha = .5)+
  scale_color_viridis_d(option = "magma", end = .9)+
  facet_wrap(~PARAM, scales = "free_y")
```
### outliers
 - Cd has one major outlier at ~40000
 Cu - has 1 high outlier > 2e5
 Hg - 1 outlier > 5000
 Pb - a whole cluster of outliers >100000

### convert ug/kg to mg/kg
```{r}
ices_mytilus.edulis_cd.pb.hg.cu$Value.mgperkg <- ices_mytilus.edulis_cd.pb.hg.cu$Value / 1000
```

### drop unneeded columns
```{r}
ices_mytilus.edulis_cd.pb.hg.cu.select <- ices_mytilus.edulis_cd.pb.hg.cu %>%
  select(Country, MYEAR:Species,NOINP, MATRX, Value.mgperkg,PARAM:QFLAG, DETLI:UNCRT)%>%
  mutate(DETLI = DETLI/1000)%>%
  rename(DETLI.mgperkg = DETLI)
```

### censored data...
```{r}
# Quality flags - many were censored at the detection or limit of quantification
table(ices_mytilus.edulis_cd.pb.hg.cu.select$QFLAG)
summary(ices_mytilus.edulis_cd.pb.hg.cu.select$DETLI.mgperkg)
#just something to keep in mind

```



